An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-07-15 paper/main.tex
praes05 upload
kurz vor Klausur

\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{float}
\usepackage{derivative}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}

\usepackage[style=nejm,
	citestyle=numeric-comp,
	sorting=none,
	doi=false,
	backend=bibtex]{biblatex}
\addbibresource{literature.bib}

@@

\begin{abstract}
	TODO
\end{abstract}

@@

Training is conducted via backpropagation, where gradients of a loss function
\( L(\hat{y}, y) \) with respect to the model parameters are computed. For
example, using mean squared error:
% \[
% L(\hat{y}, y) = (\hat{y} - y)^2, \qquad \pdv{L}{\omega_i} = 2(\hat{y} - y) \cdot \odv{\sigma}{a} \cdot x_i.
% \]

@@

Memory and compute requirements during training are substantial, as gradients,
activations, and parameters must be stored and updated iteratively. In
contrast, inference requires only a forward pass, making it more amenable to
optimization via precision reduction.

@@

To reduce memory usage and computational load, precision-aware techniques such
as quantization and mixed-precision training are increasingly employed. Instead
of using 32-bit or 64-bit floating-point formats, parameters can be stored in
16-bit or even 8-bit fixed-point representations.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{figures/float.png}
	\caption{Illustration of standard floating-point format.}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{figures/binary-formats.png}
	\caption{Comparison of binary formats used in neural network computations.}
\end{figure}

@@

We trained a simple fully connected neural network on the MNIST dataset,
consisting of 60,000 training and 10,000 test grayscale images of handwritten
digits (28×28 pixels). Model training was performed under varying parameter
bitwidths, with cross-validation to ensure robustness.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\textwidth]{figures/mnist.png}
	\caption{Sample images from the MNIST dataset.}
\end{figure}

\section{Results}

\subsection*{Accuracy Across Bitwidths}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\textwidth]{figures/trainingAndInference.png}
	\caption{Model accuracy during training and inference under different precision settings.}
\end{figure}

\subsection*{Training and Inference Time}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\textwidth]{figures/times.png}
	\caption{Execution times for various precision configurations.}
\end{figure}

\subsection*{Quantization After Training}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\textwidth]{figures/inferenceOnly.png}
	\caption{Inference accuracy using float64-trained weights followed by quantization.}
\end{figure}

\subsection*{Automatic Mixed Precision and Gradient Clipping}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\textwidth]{figures/gradclip_32.png}
	\caption{Inference accuracy using float32-trained weights with AMP and gradient clipping.}
\end{figure}

\subsection*{Effect of Decimal Rounding}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\textwidth]{figures/118_accuracy.png}
	\caption{Accuracy vs.\ number of decimal digits retained in weights.}
\end{figure}

@@

Our results indicate that networks trained and evaluated with as low as 16-bit
precision can achieve comparable performance to their high-precision
counterparts. Moreover, rounding pretrained weights to just 3 decimal digits
preserves full accuracy, opening the door to aggressive post-training
quantization.

This is particularly relevant for edge devices and real-time applications where
memory and compute constraints are paramount.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{float}
\usepackage{derivative}
\usepackage{subcaption}

\usepackage[
	style=nejm,
	citestyle=numeric-comp,
	sorting=none,
	doi=false,
	backend=bibtex]{biblatex}
\addbibresource{literature.bib}

@@

\begin{abstract}
	We study how the numerical precision of weights, activations, and gradients in neural networks can be reduced without sacrificing too much of the classification performance. Using MNIST as a dataset, we benchmark uniform 16–64-bit quantisation, mixed-precision formats, and post-training decimal rounding. The results reveal that (i) 16-bit end-to-end training preserves full accuracy for the MNIST dataset and the network used, (ii) three-decimal-place weight rounding while keeping 32-bit precision incurs minimal accuracy loss. These findings verify limits for memory- and energy-constrained usecases like inference in mobile devices.
\end{abstract}

@@

Training is conducted via backpropagation, where gradients of a loss function
\( L(\hat{y}, y) \) with respect to the model parameters are computed. For
example, using the mean squared error:
% \[
% L(\hat{y}, y) = (\hat{y} - y)^2, \qquad \pdv{L}{\omega_i} = 2(\hat{y} - y) \cdot \odv{\sigma}{a} \cdot x_i.
% \]

@@

Memory and computational requirements during training are substantial, as
gradients, activations, and parameters must be stored and updated iteratively.
In contrast, inference requires only a forward pass, making it more amenable to
optimization via precision reduction.

@@

To reduce memory usage and computational load, precision-aware techniques such
as quantization and mixed-precision training are increasingly employed. Instead
of using 32-bit or 64-bit floating-point formats, parameters can be stored in
16-bit or even 8-bit fixed-point representations.
Figure~\ref{fig:floatingPointFormat} depicts the general floating point format
and terminology of significand, base, and exponent.
Tables~\ref{tab:binary-bits} and~\ref{tab:binary-precision} provide an overview
of the different floating point types and their respective characteristics.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{figures/float.png}
	\caption{Illustration of standard floating-point format.}\label{fig:floatingPointFormat}
\end{figure}

\begin{table}[H]
	\centering
	\begin{tabular}{@{}lcccc@{}}
		\toprule
		\textbf{Type} & Sign & Exponent & Significand & Total \\
		\midrule
		Half  (FP16)  & 1    & 5        & 10          & 16    \\
		Single (FP32) & 1    & 8        & 23          & 32    \\
		Double (FP64) & 1    & 11       & 52          & 64    \\
		\bottomrule
	\end{tabular}
	\caption{Field widths (in bits) of common IEEE754 formats (adapted from~\cite{fparithmetic}).}\label{tab:binary-bits}
\end{table}

\begin{table}[H]
	\centering
	\begin{tabular}{@{}lccc@{}}
		\toprule
		\textbf{Type} & Exponent bias & Precision (bits) & Decimal digits \\
		\midrule
		Half  (FP16)  & 15            & 11               & $\sim$3.3      \\
		Single (FP32) & 127           & 24               & $\sim$7.2      \\
		Double (FP64) & 1023          & 53               & $\sim$15.9     \\
		\bottomrule
	\end{tabular}
	\caption{Range and precision characteristics of the same formats (adapted from~\cite{fparithmetic}).}\label{tab:binary-precision}
\end{table}

@@

We trained a simple fully connected neural network on the MNIST dataset,
consisting of 60,000 training and 10,000 test grayscale images of handwritten
digits (28×28 pixels, see figure~\ref{fig:mnistSample}). Model training was
performed under varying parameter bitwidths, and 5-fold cross-validation to
evaluate the model performance. The structure of the MLP is depicted in
figure~\ref{fig:mlpStructure}.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/mnist.png}
	\caption{Sample images from the MNIST dataset.}\label{fig:mnistSample}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/mlpStruc.png}
	\caption{MLP structure used in the experiments.}\label{fig:mlpStructure}
\end{figure}

\section{Results}
In the first experiment, every floating-point tensor that is used inside the
training loop was converted to a lower precision data type. This therefore
includes the weights and biases, activations, gradients, optimizer states, etc.
The types tried were float64, float32, and float16.

\subsection*{Accuracy Across Bitwidths}
Figure~\ref{fig:modelAccTrainAndInf} shows how the accuracy changed after the
models were trained on the different precision settings. Keep in mind that the
y-axis is not starting from zero, so the absolute height differences are
amplified visually, though the delta is only around $.1$ percent. Re-training
the models and re-evaluating them has also lead to the accuracies being
``reversed'', meaning the float64 model had the least accuracy. It is safe to
assume that the model accuracy has neither decreased, nor increased.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/trainingAndInference.png}
	\caption{Model accuracy evaluated after training with the different precisions.}\label{fig:modelAccTrainAndInf}
\end{figure}

\subsection*{Training and Inference Time}
Also analyzed were the times needed for both the training and inference.
Figure~\ref{fig:modelEvalTrainTimes} depicts the times measured in the training
loop and later on at the inference stage. Again, the times are almost identical
in both regards, and neither a real speed-up nor a slow-down has been measured.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/times.png}
	\caption{Training and evaluation times for the different precisions.}\label{fig:modelEvalTrainTimes}
\end{figure}

\subsection*{Inference-Only Testing}
This third experiment was conducted to test a possible post-training precision
reduction. For this, the model was trained on the highest precision (float64)
and then quantized at the inference stage.
Figure~\ref{fig:inferenceOnlyTesting} shows that the accuracy of the baseline
(float64 training AND inference) is maintained as the precision is decreased at
the later stages. In contrast to the \textit{Accuracy Across Bitwidths}
experiment, there are no fluctuations at all because for the lower precisions,
there was no new model trained.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/inferenceOnly.png}
	\caption{Inference accuracy using float64-trained weights followed by quantization.}\label{fig:inferenceOnlyTesting}
\end{figure}

\subsection*{Automatic Mixed Precision and Gradient Clipping}
As a next experiment, \dots %Sayed, also use the other images?
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/gradclip_32.png}
	\caption{Inference accuracy using float32-trained weights with AMP and gradient clipping.}
\end{figure}

\subsection*{Effect of Decimal Rounding}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_accuracy.png}
	\caption{Accuracy vs.\ number of decimal digits retained in weights.}
\end{figure}

\section{Summary}
\begin{itemize}
	\item \textbf{16-bit is enough for this classification task}: End-to-end float16 preserved the accuracy.
	\item \textbf{Three decimals suffice}: Storing weights down to fractional length $3$ caused no visible accuracy loss.
\end{itemize}

@@

The results of the first experiments have shown that either the selected task
(MNIST classification) was not complex enough to be affected by the decrease in
accuracy, or the accuracy would need to be decreased even further than float16.

Otherwise, the results generally indicate that networks trained and evaluated
with as low as 16-bit precision can achieve comparable performance to their
high-precision counterparts. Moreover, rounding pretrained weights to just 3
decimal digits preserves full accuracy, opening the door to aggressive
post-training quantization.

This is particularly relevant for edge devices and real-time applications where
memory and compute constraints are more relevant.
